First steps in synthetic algebraic geometry

### 2017-02-06 slides-warwick2017.tex
Slides almost finished

\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{mathtools}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}
\hypersetup{colorlinks=true}

@@

\usefonttheme{serif}
\usepackage[T1]{fontenc}
\usepackage{libertine}
%\usepackage{mathpazo}

\newcommand{\A}{\mathcal{A}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\DeclareMathOperator{\Spec}{Spec}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1_S}}}
\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}

@@

\begin{frame}\frametitle{A curious property of the structure sheaf}
  Let~$X$ be a scheme. Internally to~$\Sh(X)$,
  \begin{center}
    \hil{any non-invertible element of~$\boldsymbol{\O_X}$ is nilpotent.}
  \end{center}

@@

@@

\begin{frame}\frametitle{Synthetic algebraic geometry}
  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
  using either
  \begin{itemize}
    \item locally ringed spaces
    \begin{multline*}
      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^2+Y^2-Z^2)$} + {} \\
      \text{Zariski topology} + \text{structure sheaf}
    \end{multline*}
    \item or Grothendieck's functor-of-points approach, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
    \[ A \longmapsto \{ (x,y,z) \in A^ 3 \,|\, x^2+y^2-z^2=0 \} \]
  \end{itemize}

  \hil{Synthetic approach}: model schemes \hil{directly as sets} in a certain
  nonclassical set theory.
  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^2+y^2-z^2=0 \} \]
\end{frame}

@@

\begin{frame}\frametitle{Some internal constructions}
  \begin{itemize}
  \item $\mathbb{P}^n_S \defeq \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times.$

  \item $\RelSpec_S A \defeq \Hom_{\mathrm{Alg}(\affl)}(A, \affl).$

  \item $TX = \Hom(\Delta, \ull{X})$,
  where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.
  \end{itemize}
\end{frame}

@@

Mention:
* generic freeness

>>>>

%\usepackage{mathtools}
\newcommand{\vcentcolon}{:}
\usepackage{booktabs}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{multicol}
\usepackage{tabto}
\usepackage{xstring}
\usepackage{soul}\setul{0.3ex}{}
\usepackage[all]{xy}
\xyoption{rotate}
\usepackage{tikz}
\usetikzlibrary{calc,shapes.callouts,shapes.arrows}
\hypersetup{colorlinks=true}

@@

\usefonttheme{serif}
%\usepackage[T1]{fontenc}
%\usepackage{libertine}

\newcommand{\A}{\mathcal{A}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\DeclareMathOperator{\Spec}{Spec}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\vcentcolon\!\Longleftrightarrow}

@@

\begin{frame}\frametitle{\mbox{A curious property of the structure sheaf}}
  Let~$X$ be a scheme. Internally to~$\Sh(X)$,
  \begin{center}
    \hil{any non-invertible element of~$\boldsymbol{\O_X}$ is nilpotent.}
  \end{center}

@@

\begin{frame}\frametitle{Generic freeness}
  % XXX: enlarge diagram
  \vspace*{-1em}
  \[ \xymatrix{
    & \ar@{-}[d]^{\substack{\text{finitely}\\\text{generated}}} M \\
    R \ar[r]_{\text{of finite type}} & S
  } \]
  If $R$ is reduced ($x^n = 0 \Rightarrow x = 0$), there is a \hil{dense open}~$U \subseteq \Spec R$ such that for any $f \in R$ with~$D(f)
  \subseteq U$,
  \begin{itemize}
    \item $S[f^{-1}]$ and $M[f^{-1}]$ are free modules over $R[f^{-1}]$,
    \item $R[f^{-1}] \to S[f^{-1}]$ is of finite presentation, and
    \item $M[f^{-1}]$ is finitely presented as a module over~$S[f^{-1}]$.
  \end{itemize}

  For a trivial proof, employ~$\Sh(\Spec R)$ and exploit that
  \begin{itemize}
    \item $\O_{\Spec R}$ is a \hil{field}: $\neg(\text{$x$ invertible}) \Rightarrow x = 0$,
    \item $\O_{\Spec R}$ and $\O_{\Spec R}[X_1,\ldots,X_n]$ are \hil{Noetherian}
    in the sense that any ideal is \hil{not not} finitely generated.
  \end{itemize}
\end{frame}

@@

\begin{frame}\frametitle{Synthetic algebraic geometry}
  Usual approach to algebraic geometry: \hil{layer schemes above ordinary set theory}
  using either
  \begin{itemize}
    \item locally ringed spaces
    \begin{multline*}
      \text{set of prime ideals of~$\ZZ[X,Y,Z]/(X^3+Y^3-Z^3)$} + {} \\
      \text{Zariski topology} + \text{structure sheaf}
    \end{multline*}
    \item or Grothendieck's functor-of-points approach, where a scheme is a functor~$\mathrm{Ring} \to \mathrm{Set}$.
    \[ A \longmapsto \{ (x,y,z) \in A^ 3 \,|\, x^3+y^3-z^3=0 \} \]
  \end{itemize}

  \hil{Synthetic approach}: model schemes \hil{directly as sets} in a certain
  nonclassical set theory.
  \[ \{ (x,y,z) \in (\affl)^3 \,|\, x^3+y^3-z^3=0 \} \]
\end{frame}

@@

\begin{frame}\frametitle{Some internal constructions}
  \begin{itemize}
  \item $\mathbb{P}^n_S = \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\,
    x_0 \neq 0 \vee \cdots \vee x_n \neq 0 \}/(\affl)^\times.$

  \item $\RelSpec_S A = \Hom_{\mathrm{Alg}(\affl)}(A, \affl) =
  \text{set of $\affl$-rational points of $A$}.$

  \item $TX = \Hom(\Delta, X)$,
  where~$\Delta = \{ \varepsilon : \affl \,|\, \varepsilon^2 = 0 \}$.

  \item An $\affl$-module~$E$ is quasicoherent if and only if
  \[ E \otimes_\affl A \longrightarrow \Hom(\Spec A, E) \]
  is an isomorphism for all finitely presented $\affl$-algebras $A$.

  \item In particular, any map $\affl \to \affl$ is given by a polynomial.

  \item A subset $U \subseteq X$ is open if and only if for any $x : X$
  there exists an ideal $J \subseteq \affl$ such that $\affl/J$ is quasicoherent and
  \[ x \in U \quad\Longleftrightarrow\quad J = (1). \]

  \item A subset $U \subseteq X$ is qc-open if and only if for any $x : X$
  there exist $f_1,\ldots,f_n \subseteq \affl$ such that
  \[ x \in U \quad\longleftrightarrow\quad \exists i\_ \text{$f_i$ is invertible}. \]
  \end{itemize}
\end{frame}

@@
